### 2022-09-12 AISTATS2023/paper.tex
updated method description

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML models that not only predict
accurately but also explain their predictions. Therefore, Explainable
AI (XAI), the field that provides interpretations for the prediction
mechanism of complex black-box models, has increased interest. XAI
literature distinguishes between local and global interpretation
methods~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to know which features are
significant (feature importance) or whether a particular feature has a
positive or negative effect on the output (feature effect), they
should opt for a global explainability technique. Aggregating the
individual explanations for producing a concise global one comes at a
cost. Under strong feature interactions, the global explanation may
obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist
under the hood; a phenomenon called aggregation
bias~\citep{mehrabi2021survey}.

Feature effect\citep{Gromping2020MAEP} forms a fundamental category of
global explainability methods, isolating a single feature's average
impact on the output. Feature effect methods suffer from aggregation
bias because the rationale behind the average effect might be
unclear. For example, a feature with zero average effect may indicate
that the feature has no effect on the output or, contrarily, it has a
highly positive effect in some cases and a highly negative in
others. There are three widely-used feature effect methods; Partial
Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is the case in most ML setups. Therefore, ALE has
been established as the state-of-the-art feature effect method.

However, ALE faces two crucial drawbacks. First, it does not provide a
way to inform the user about potential heterogeneous effects that are
hidden behind the average effect. In contrast, in the case of PDP, the
heterogeneous effects can be spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
ALE requires an additional step, where the axis of the feature of
interest is split in \(K\) fixed-size non-overlapping intervals, where
\(K\) is a hyperparameter provided by the user. So far, this splitting
is done in a blind way, i.e., the user does not have an indication
about the parameter K, which often leads to inconsistent explanations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the global explanation. The uncertainty
of the global explanation expresses how certain we are that the global
(expected) explanation is valid if applied to an instance drawn at
random and informs the user about the level of heterogeneous effects
behind the expected explanation. The probabilistic extension completes
ALE, as ICE plots complement PDP, for revealing the heterogeneous
effects. Our method also transforms the axis-splitting step into an
unsupervised clustering problem, i.e., we search for the optimal
splitting given the instances of the training set. We, therefore,
relieve the user from the obligation of defining a parameter \(K\) by
providing an automated method for finding the optimal bin splitting to
robustly estimate (a) the global (expected) effect and (b) the
uncertainty of the explanation, from the limited samples of the
training set.

@@

\begin{itemize}
\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
  quantifies the uncertainty of the global explanation, i.e.~the level
  of heterogeneous effects hidden behind the global explanation.
\item We provide an algorithm that automatically computes the optimal
  axis-splitting for robustly estimating the the global effect and the
  uncertainty.
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity that is the lower-bound of the objective.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

@@

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), whereas to simple variables with plain lowercase
\( x \). Bold denotes a vector; \( \xb \) for simple variables or
\(\mathbf{X}\) for rvs. Often, we partition the input vector
\(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} As stated at the Introduction, the
three well-known feature effect methods are: PDPs, MPs and ALE. PDPs
formulate the feature effect of the \(s\)-th attribute as an
expectation over the marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects, which are defined as the
change on the output, i.e.
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
corelated input features. In these cases, PDPs integrate over
unrealistic instances, due to the use of the marginal distribution
\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
impute the combined effect of sets of features to a single
feature. ALE manages to resolve both issues, and is therefore the only
trustable method in cases of correlated features.

@@

Feature effect methods answer the question \textit{what happens
  (effect) to the output, if I increase/decrease the value of a
  specific feature}. Having answered the question above, it comes
naturally to also wonder \textit{how certain we are about the change
  on the output}. For this reason, a lot of interest is given lately
for quantifying the level of uncertainty, along with the expected
effect. The level of uncertainty is mostly quantified by measuring the
existence of heterogeneous effects, i.e. whether there are local
explanations that deviate from the expected global effect. ICE and
d-ICE plots provide a visual understanding of the heterogeneous
effects on top of PDPs. Another approach targets on grouping the
heterogeneous effects, e.g., allocating ICE plots in homogeneous
clusters, by dividing the input space. Some other approaches, like
H-Statistic, Greenwel, move a step behind and try to quantify the
level of interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches face two pathogenies; They
either do not quantify the uncertainty of the feature effect directly
or they are based on PDPs, and, therefore, they are subject to the
failure modes of PDPs in cases of correlated features. To the best of
our knowledge, no method so far targets on quantify the uncertainty of
the feature effect as it is modelled by ALE.

\paragraph{Cluster Instances with homogeneous effects.}

In real ML scenarios, the expected feature effect and the uncertainty
are estimated from the limited instances of the training set. ALE
approximation requires an additional step, where the axis of the
\(s\)-th feature is split into a sequence of non-overlaping bins and a
single effect (expectation and uncertainty) is computed from the
population of instances that lie inside each
bin. \citep{apley2020visualizing} proposed estimating the local
effects in each bin by evaluating the black box-funtion at the bin
limits:

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}

We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). In
contrast, (cite) proposed the Differential ALE (DALE) estimation for
quantifying the local effects on the training-set instances, instead
of the bin limits:

\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, it allows the recomputation
of the accumulated effect with different bin-splitting in near-zero
computational overhead. However, none of the approximations above
deals with the crucial problem of the optimal bin-spliting. They
partition the axis in \(K\) equally-sized bins without considering the
properties of the underlying data, which can lead to erroneous
approximations.

Instead, we propose treating the axis-spliting step as an unsupervised
clustering problem. The objective of the clustering problem should
fulfill in the best way to contradictory ojectives. First, secure
robust estimations of the expected effect and the uncertainty inside
each bin given the limited instances of the training set and, second,
create bins with as homogeneous local effects as possible, for not
losing fine-grain resolution feature effects due to wide bins.

@@

The NAME method extends the traditional ALE method with a component
for uncertainty quantification and improves the ALE estimation by
automatically discovering the optimal bin-splitting. In
Section~\ref{sec:NAME-definition} we define the component for the
uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
NAME from the limited samples of the training set and we make an
important proof about the aggregated variance defined over a bin. In
Section~\ref{sec:interval-spliting}, we define and solve the problem
of optimal bin-splitting. Finally, in Section~\ref{sec:visualization},
we illustrate the appropriate visualization of NAME for facilitating
its interpretation by a non-expert and we discuss important aspects of
the method

@@

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The feature effect at \(x_s\) is the accumulation of the averaged
local effects from \(x_{s, min}\) until \(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). As described at the Introduction, limiting the explanation to the
exepected value level does not shed light to possible heterogeneous
effects behind the averaged explanation. Therefore, we model the
uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
feature effect at \(x_s\) is the accumulation of the averaged local
effects from \(x_{min}\) until \(x_s\), as shown in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until \(x_s\), as the integral of
the variances of the local effects:

@@

@@

\noindent
TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)

% Furthermore, we automate the step of splitting the axis into non-overlapping intervals. The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set. Therefore, there is an implicit trade-off behind the formation of bins. Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance). On the other hand, each bin should include points with similar local effects. Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function. We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation. Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.

>>>>

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
ther other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{Herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visulalize the impact of a single feature on the
output.~\footnote{FE methods also isolate the effect of a pair of
  features to the output. Combinations of more than two features are
  not usual, because they encounter, among others, visualization
  difficulties.} FE methods suffer from aggregation bias because,
often, the rationale behind the average effect might be unclear. For
example, a feature with zero average effect may indicate that the
feature has no effect on the output or, contrarily, it has a highly
positive effect in some cases and a highly negative in others. There
are three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML
problems. Therefore, ALE has been established as the state-of-the-art
FE method.

However, ALE faces two crucial drawbacks. First, it does not provide a
way to inform the user about potential heterogeneous effects that are
hidden behind the average effect. In contrast, in the case of PDP, the
heterogeneous effects can be spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
ALE requires an additional step where the axis of the feature of
interest is split in \(K\) fixed-size non-overlapping intervals. So
far, the user is asked to provide a value for the parameter \(K\)
blindly, i.e without an indication about whether a big or a small
value could be more appropriate, which often leads to unstable
explanations. For convenience, in the rest of the paper we will refer
to this step using the term bin-splitting problem.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the global explanation. The uncertainty
of the global explanation expresses how certain we are that the global
(averaged) explanation is valid if applied to an instance drawn at
random. The probabilistic extension completes ALE, as ICE plots
complement PDP, for revealing the heterogeneous effects. Furthermore,
we transform the bin-splitting step into a data-driven clustering
problem, i.e., we search for the optimal splitting given available
instances of the training set. We, also, present a
computationally-grounded algorithm for finding the optimal
solution.

@@

\begin{itemize}
\item We introduce NAME, an extension of ALE that quantifies the
  uncertainty of the global explanation, i.e.~the level of
  heterogeneous effects hidden behind the global explanation.
\item We present an algorithm that automatically computes the optimal
  bin-splitting
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

@@

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), whereas to simple variables with plain lowercase
\( x \). Bold denotes a vector; \( \xb \) for simple variables or
\(\mathbf{X}\) for rvs. Often, we partition the input vector
\(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
FE of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the FE method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} The three well-known feature
effect methods are: PDP, MP and ALE. PDP formulates the FE
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects:

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
corelated input features. In these cases, PDP integrates over
unrealistic instances, due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the only trustable
FE method in cases of correlated features.

@@

FE methods answer the question \textit{what is expected to happen to
  the output (expected effect), if the value of a specific feature is
  increased/decreased}. Having an answer to the question above, it
comes naturally to also wonder \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty along with the expected effect has attracted a lot
of interest. The level of uncertainty is mostly quantified by
measuring the existence of heterogeneous effects, i.e. whether there
are local explanations that deviate from the expected global
effect. ICE and d-ICE plots provide a visual understanding of the
heterogeneous effects on top of PDPs. Another approach targets on
grouping the heterogeneous effects, e.g., allocating ICE plots in
homogeneous clusters, by dividing the input
space.\citep{molnar2020model} Some other approaches, like H-Statistic,
Greenwel, move a step behind and try to quantify the level of
interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches face two pathogenies; They
either do not quantify the uncertainty of the FE directly or they are
based on PDPs, and, therefore, they are subject to the failure modes
of PDPs in cases of correlated features. To the best of our knowledge,
no method so far targets on quantifying the uncertainty of ALE.

\paragraph{Bin-Splitting for ALE estimation.}

In real ML scenarios, the expected FE and the uncertainty are
estimated from the limited instances of the training set.
\citep{apley2020visualizing} proposed estimating the local effects in
each bin by evaluating the black box-funtion at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE) that
exploits differentiaton for computing the local effects on the instances
training-set, instead of the bin limits:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost. However, none of the approximations above deals
with the crucial problem of bin-spliting. As indicated
by~\citep{molnar2022}, in ALE the effects are computed per interval
(region) and the interpretation of the effect can only be
local.

@@

In Section~\ref{sec:NAME-definition} we define the component for the
uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
the average effect and the uncertainty from the limited samples of the
training set and we make an important proof about the aggregated
variance defined over a bin. In Section~\ref{sec:interval-spliting},
we define and solve the problem of optimal bin-splitting. Finally, in
Section~\ref{sec:visualization}, we illustrate the appropriate
visualization of NAME for facilitating its interpretation by a
non-expert and we discuss important aspects of the method

@@

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The FE at \(x_s\) is the accumulation of the averaged
local effects from \(x_{s, min}\) until \(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). As described at the Introduction, limiting the explanation to the
exepected value level does not shed light to possible heterogeneous
effects behind the averaged explanation. Therefore, we model the
uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
FE at \(x_s\) is the accumulation of the averaged local
effects from \(x_{min}\) until \(x_s\), as shown in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until \(x_s\), as the integral of
the variances of the local effects:

@@

% Instead, we propose treating the axis-spliting step as an unsupervised
% clustering problem. The objective of the clustering problem should
% fulfill in the best way to contradictory ojectives. First, secure
% robust estimations of the expected effect and the uncertainty inside
% each bin given the limited instances of the training set and, second,
% create bins with as homogeneous local effects as possible, for not
% losing fine-grain resolution FEs due to wide bins.

@@

\noindent

\begin{itemize}
\item Computational complexity
\end{itemize}

% Furthermore, we automate the step of splitting the axis into non-overlapping intervals. The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set. Therefore, there is an implicit trade-off behind the formation of bins. Each bin must include enough instances for a robust estimation of the bin FE (expected value), and the uncertainty of the explanation (variance). On the other hand, each bin should include points with similar local effects. Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function. We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation. Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.
